Synthesising Complex Termination Arguments

### 2014-06-16 papers/termination/paper.tex
Change results table
revised intro+motivation+other things

Proving program termination is typically done by finding a well-founded \emph{ranking function}
for the program states. When doing so, the majority of the existing termination provers diverge 
in one important aspect from the execution
of a program on a computer:
bit-vectors and floats 
are treated as mathematical integers and reals, respectively \cite{}.
This approximation gives rise to 
{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

For the case of bit-vectors (machine-level integers), their abstraction  to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
incomparable behaviours with respect to termination:
%The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
\begin{itemize}
\item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}
The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{}). 
\item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
 \begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
\end{itemize}

A scenario similar to the one for bit-vectors happens when floats are abstracted to unbounded reals by termination provers \cite{}.
Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated next:
\begin{itemize}
\item This loop does not terminate for reals, but does for floats.
\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}

\item The following loop terminates for reals, but does not for floats.
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\end{itemize}   
\todo{explain the reasons for non-termination with figures.}

Another limitation of the state of the art in the computation of ranking functions is the fact that  most techniques assume \emph{restrictive transition relations} and are only able to compute
\emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.

Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
reachable after leaving a previous loop.

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
(lexicographic) linear and non-linear ones, 
for programs with both bit-vectors and floats. Our technique can handle programs with non-linear operations, e.g. logical and.
For this purpose, we rephrase the termination problem as a second-order satisfaction problem
(indeed, a program synthesis problem). 
As a direct consequence, our technique is semantic in nature and does not
dependent on the size of the analysed program or its number of variables. 
However, its efficiency does depend on the Kolmogorov
 complexity of the computed ranking functions. We show that the probability of
a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).

By phrasing the termination problem as a single rank function synthesis problem, we observe that
we can make the termination problem as hard as finding ranking functions, rather than as hard as
solving a difficult ``double reachability'' problem (cf. Terminator).

\begin{figure}
 \begin{tabular}{|ll||c|c|}
 \hline
  & & \multicolumn{2}{c|}{Program} \\
  & & Linear & Non-Linear \\
  \hline
  \hline
  \multirow{2}{*}{Ranking Functions} & Linear & \checkmark \xmark & \checkmark \\
   & Non-Linear & \checkmark & \checkmark \\
   \hline
 \end{tabular}

 \caption{Legend: \xmark = state-of-the art can handle, \checkmark = we can handle\label{fig:handletable}}
\end{figure}

@@

The main contributions of our paper are highlighted below:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We present a formulation that allows us to prove termination without using a fancy safety prover (because
we don't do a DWF proof).  In particular,
we only need a bounded model checker that performs a single unwinding of a loop.
\item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
multiple loops.
\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
\end{itemize}

%completeness without disj well-foundedness
%. only non-strict inequalities can be transformed using Farkas’ Lemma

@@

\section{Related Work}
\subsection{Termination analysis}
The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.

@@

Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
%Terminator incrementally constructs the union of ranking relations.
%Given that the this union is not well-founded (it is only disjunctively well-founded), 
%Terminator must prove that the binary reachability relation of the program (the transitive closure of the transition relation restricted to reachable states) 
%is contained in the union of ranking relations.
%Thus, this approach requires an evolved binary reachability, and, implicitly, the assistance of a safety prover.
In particular, their technique incrementally constructs the union of ranking relations.
Given that the this union is not well-founded (it is only disjunctively well-founded),
they must show that, for every sequence, one of the ranking functions decreases
between the pre- and post-state. %In other words: it 
%must find all pairs of states reachable from the program's initial state 
%such that one is reachable from the other and it must show 
%that the value of one of the ranking functions decreases from s1 to s2. 
Consequently, most of the analysis time is spent in reachability analysis that is to check if the current
transition invariant reached a fixed point.

@@

@@

\section{Motivational Examples} 
We next discus a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
frequently in low-level code and are not handled by existing techniques.

@@

\subsection{A non-linear program with a linear ranking function.}
The first example we consider is a program with non-linear operations: 
\begin{lstlisting}
while (x > 0)
 x = (x - 1) & x;
\end{lstlisting}

@@

\subsection{A linear program with a non-linear ranking function.}
\begin{lstlisting}
while (x != 0)
 x = -x / 2;
\end{lstlisting}

@@

\subsection{A linear program with lexicografic non-linear function.}
\begin{lstlisting}
while ( q >= 0 ) {
  q = q - y;
  y = y + 1;
}
\end{lstlisting}

Every execution of the above program can be partitioned into two phases: in the first phase $y$ increases
until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
%Depending on the initial values of y and q, either phase might be skipped
%altogether.
\cite{DBLP:conf/tacas/LeikeH14} proposes using a linear ranking template in order to obtain a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is
ranked by an affine-linear function, and ends when this function becomes non-positive.

\subsection{Lexicographic ranking function with strict inequalities.}
Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
[2,5,6,7,14,20,24,25

\section{Preliminaries}
\subsection{Termination and Ranking Functions}
A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  We can prove that $T$ is
unconditionally terminating by finding an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect $T$.  That is
to say:
$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$

@@

\subsection{Machine Arithmetic and Bitvectors}
Computers tend to have fixed width integers.  For a machine with $k$-bit words,
all arithmetic is done modulo $2^k$.  This means that a program's termination
behaviour can depend strongly on whether it is interpreted as operating over
mathematical integers, or over fixed-width machine words.  We can however
observe that if no arithmetic operation \emph{overflows} (i.e. does not result
in a value greater than $2^k$), fixed-width and integer arithmetic coincide.

@@

>>>>

Proving program termination is typically done by finding a \emph{ranking function}
for the program states, i.e. a map from the program's state space to a well-ordered set.
As this definition of ranking function is very general, research is often limited to some
convenient and tractable form of ranking functions, most frequently \emph{linear ranking functions} (see Section~\ref{sec:ranking.functions}). 
In doing so, an analysis ensures that such a ranking function will be found is one  exists, 
while failing to prove termination for terminating programs whose ranking functions fall short of the considered restriction.

Moreover, the majority of the existing termination provers diverge 
in one important aspect from the execution of a program on a computer:
bit-vectors and floats 
are treated as mathematical integers and reals, respectively.
This approximation gives rise to 
{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

Figure~\ref{fig:handletable} presents a summary of existing termination analyses with indeed
most of the work concentrated on finding \emph{linear ranking functions for linear programs over integers}.
Next, we will look in more detail at the implications of the assumptions made by such techniques.\\

\noindent {\bf Bit-vectors (machine-level integers) vs. mathematical integers.} The abstraction of bit-vectors to mathematical integers 
ignores the wrap-around behaviour caused by under/over-flows in bit-vector arithmetic, resulting in %, it gives rise to 
%{\em unsound} and {\em incomplete} analyses, which is rather surprising given that bit-vectors are ubiquitous in computer systems. 
incomparable behaviours with respect to termination:
%The differences between the semantics of unbounded mathematical integers and that of machine integers result in 
\begin{itemize}
\item Programs that terminate for integers may \emph{not} terminate for bit-vectors. For illustration, consider the following loop:
\begin{lstlisting}[language=C]
while (x > 0) x -= 2;
\end{lstlisting}
The loop always terminates for unbounded integers as the value of \texttt{x} does eventually become negative, 
whereas, with bit-vector arithmetic, \texttt{x} under-flows before becoming negative and goes back to being positive causing the loop to never terminate.
According to the C11 standard for the C programming language, a signed under- or over-flow yields undefined behaviour.
However, compilers generally treat signed under- and over-flows using the
wrap-around behaviour (see Section~\ref{sec:machine.arith}). 
\item Similarly, programs that terminate for bit-vectors may \emph{not} terminate for integers. One such situation is illustrated next:  
 \begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since the loop condition stays always true for any initial \texttt{x} at least 1.
\end{itemize}

\noindent {\bf Floats vs. reals.} A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated next:
\begin{itemize}
\item This loop does not terminate for reals, but does for floats.
\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}

\item The following loop terminates for reals, but does not for floats.
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\end{itemize}   
\todo{explain the reasons for non-termination with figures.}\\

\noindent {\bf Linear programs and linear ranking functions.} As visible in Figure~\ref{fig:handletable}, 
most termination techniques assume \emph{restrictive transition relations}, e.g. linear programs,  and are only able to compute
\emph{linear ranking functions} \cite{}. To better understand this restriction, we computed the probability of a random terminating program having a linear ranking
function (see Section ?). This probability proved to be very low, indicating that the linearity assumption for the ranking functions is indeed prohibitive.

We propose a general framework that does not assume the existence of linear ranking functions and can uniformly compute
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
For this purpose, we rephrase the termination problem as a second-order satisfaction problem
(indeed, a program synthesis problem). 
As a direct consequence, our technique is semantic in nature and does not
dependent on the size of the analysed program or its number of variables. 
However, its efficiency does depend on the Kolmogorov
 complexity of the computed ranking functions. We show that the probability of
a ranking function to have a low Kolmogorov complexity is higher than the probability to be linear (see Section~?).

\begin{figure*}
\centering
 \begin{tabular}{|ll||c|c|c|c|c|c|c|c|}
 \hline
  & & \multicolumn{8}{c|}{Program} \\
  & & \multicolumn{2}{c|}{Integers} & \multicolumn{2}{c|}{Reals} & \multicolumn{2}{c|}{Bit-vectors} & \multicolumn{2}{c|}{Floats} \\
  & & L & NL & L & NL & L & NL & L & NL \\
  \hline
  \hline
  \multirow{4}{*}{Ranking} & Linear lexicographic &  \cite{DBLP:conf/cav/BradleyMS05} && & &\checkmark&\checkmark&\checkmark&\checkmark\\
   & Linear non-lexicographic & \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/LeeWY12,DBLP:conf/popl/Ben-AmramG13} & & && \checkmark \cite{DBLP:conf/tacas/CookKRW10} &\checkmark \cite{DBLP:conf/tacas/CookKRW10}&\checkmark&\checkmark\\
   & Non-linear lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   & Non-linear non-lexicographic &  &  & &&\checkmark&\checkmark&\checkmark&\checkmark\\
   \hline
 \end{tabular}

 \caption{Legend: \checkmark = we can handle\label{fig:handletable}}
\end{figure*}

@@

 The main contributions of our work can be summarised as follows:
\begin{itemize}
\item We designed a technique for computing ranking functions that correctly accounts for the wrap-around behavior caused by under- and overflows in bit-vector and floating point arithmetic. To the best of our knowledge, this is the first approach able to compute ranking functions for programs handling floats. Our technique is not restricted to finding linear ranking functions, but can also compute (lexicographic) non-linear  ones. We justified the need for such a non restrictive procedure by computing the probability ... .
\item  We rephrased the termination problem as a second-order satisfaction problem and made 
use of results in genetic programming to efficiently solve it. We have also investigated the effects of genetic operators on the search space for ranking functions and computed theoretical 
bounds on the convergence time ...
\item We present a formulation that allows us to prove termination without using a fancy safety prover (because
we don't do a DWF proof).  In particular,
we only need a bounded model checker that performs a single unwinding of a loop.
\item Our technique is able to uniformly handle conditionally-terminating loops, as well as programs with
multiple loops.
\item We implemented our technique and tried it on a selection of programs handling both bit-vectors and floats.
\end{itemize}

%. only non-strict inequalities can be transformed using Farkas’ Lemma
%--Our approach to termination analysis has two distinct features over previous works
%-- a lexicographic ranking function imposes a lexicographic ordering on among the ranking function components.

Other points to touch in the introduction:
\begin{itemize}
\item disjunctive well-foundedness: By phrasing the termination problem as a single rank function synthesis problem, we observe that
we can make the termination problem as hard as finding ranking functions, rather than as hard as
solving a difficult ``double reachability'' problem (cf. Terminator).
\item The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
\item Compared to the previous approaches that mostly aim to find a particular form of ranking functions, our technique does not commit to any particular one. We will find the one that has the lowest Kolmogorov complexity.
\item Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
reachable after leaving a previous loop.

\end{itemize}

@@

\section{Related Work}
\subsection{Termination analysis}
Given the multitude of works on termination, we have tried to summarise them in Figure~\ref{fig:handletable}.

The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.

@@

@@

Linear ranking functions supported by inductive linear invariants for loops with linear guards and transitions: \cite{DBLP:conf/cav/BradleyMS05}.

\subsection{Disjunctive well-foundedness}
Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
%Terminator incrementally constructs the union of ranking relations.
%Given that the this union is not well-founded (it is only disjunctively well-founded), 
%Terminator must prove that the binary reachability relation of the program (the transitive closure of the transition relation restricted to reachable states) 
%is contained in the union of ranking relations.
%Thus, this approach requires an evolved binary reachability, and, implicitly, the assistance of a safety prover.
In particular, their technique incrementally constructs the union of ranking relations.
Given that the this union is not well-founded (it is only disjunctively well-founded),
they must show that, for every sequence, one of the ranking functions decreases
between the pre- and post-state. %In other words: it 
%must find all pairs of states reachable from the program's initial state 
%such that one is reachable from the other and it must show 
%that the value of one of the ranking functions decreases from s1 to s2. 
Consequently, most of the analysis time is spent in reachability analysis that is to check if the current
transition invariant reached a fixed point.

@@

\section{Motivational Examples} 
We next discus a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
frequently in low-level code and are not handled properly by existing techniques.

@@

\subsection{Program linearity vs. the linearity of its ranking function.}
The first examples we consider illustrate the lack of a direct correlation between the linearity of a program and that of its termination arguments.
As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
although a linear ranking function exists (see Figure~\ref{fig:handletable}).

@@

Conversely, Figure~\ref{} shows a linear program with a non-linear ranking function.

Figure~\ref{} is an example of a linear program that only admits lexicografic non-linear ranking functions.
Every execution of the above program can be partitioned into two phases: in the first phase $y$ increases
until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
%Depending on the initial values of y and q, either phase might be skipped
%altogether.
\cite{DBLP:conf/tacas/LeikeH14} proposes using a linear ranking template in order to obtain a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is
ranked by an affine-linear function, and ends when this function becomes non-positive.

\begin{lstlisting}
while (x > 0)
 x = (x - 1) & x;
\end{lstlisting}

\begin{lstlisting}
while (x != 0)
 x = -x / 2;
\end{lstlisting}

\begin{lstlisting}
while ( q >= 0 ) {
  q = q - y;
  y = y + 1;
}
\end{lstlisting}

\subsection{Differences in the termination behaviour for integers and bit-vectors.}
We have collected a number of motivational examples from other termination papers that treated bit-vectors as mathematical integers \cite{DBLP:conf/tacas/LeikeH14,DBLP:conf/tacas/CookSZ13}. 
We show that the termination arguments computed by such techniques do not directly apply when considering the bit-vector semantics.
%these programs actually exhibit different terminating behaviours for bit-vectors and integer

@@

\subsection{Multi-phase ranking functions}
%\subsection{Lexicographic ranking function with strict inequalities.}
%Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
\begin{lstlisting}
int main(void) {
  int x, y, z;

  while (x > 0 && y > 0 && z > 0) {
    if (y > x) {
      y = z;
      x = nondet();
      z = x - 1;
    } else {
      z = z - 1;
      x = nondet();
      y = x - 1;
    }
  }
}
\end{lstlisting}

\section{Preliminaries}
\subsection{Termination and Ranking Functions} \label{sec:ranking.functions}
\todo{linear program}

A transition system with state space $X$ and transition relation $T \subseteq X \times X$
is said to be \emph{unconditionally terminating} if there is no infinite sequence
$x_1, x_2, \ldots$ with $\forall i . T(x_i, x_{i+1})$.  We can prove that $T$ is
unconditionally terminating by finding an injective function $R: X \to Y$ where
$Y$ is well-founded and $R$ is monotonically decreasing with respect $T$.  That is
to say:
$$\forall x, x' . T(x, x') \Rightarrow R(x) < R(x')$$

@@

\subsection{Machine Arithmetic and Bitvectors} \label{sec:machine.arith} 
Computers tend to have fixed width integers.  For a machine with $k$-bit words,
all arithmetic is done modulo $2^k$.  This means that a program's termination
behaviour can depend strongly on whether it is interpreted as operating over
mathematical integers, or over fixed-width machine words.  We can however
observe that if no arithmetic operation \emph{overflows} (i.e. does not result
in a value greater than $2^k$), fixed-width and integer arithmetic coincide.

@@

\subsection{The Search Space and the Kolmogorov Complexity of a Ranking Function}
\begin{conjecture}
The lexicographic ordering does not influence the size of the search space. Thus, finding potentially lexicographic termination arguments is not more difficult than finding non-lexicographic ones.   
\end{conjecture}

As a consequence of the aforementioned conjecture, we do not need to commit to a specific form of the termination argument. While other approaches, e.g. \cite{DBLP:conf/tacas/LeikeH14}, 
conduct independent searches for each possible form of the ranking function with some of them futile whenever the assumed constrained ranking function does not exist, we have a single search that aims at finding the ranking function with the lowest Kolmogorov complexity.
